---
title: "A First Pass At Australian Dwelling Prices"
description: "A small reproducible look at ABS dwelling values, state mean prices and the housing policy questions they raise."
date: "2026-06-11"
categories: [housing, cost-of-living, australia, policy, abs]
author: "Aydin"
---

Housing debates have a way of turning into vibes very quickly. Prices feel high, supply feels tight, first-home buyer policy gets loud, and then everyone starts arguing from their favourite chart.

So this is a small grounding exercise: use the latest ABS dwelling-value data, look at the national dwelling stock and mean prices by state, then write down the policy questions that seem worth chasing next.

```{r}
#| label: setup
#| include: false

library(readxl)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(lubridate)
library(forcats)
library(scales)
library(knitr)

theme_set(theme_minimal(base_size = 12))

raw_dir <- file.path("data-raw", "housing")
dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)

source_urls <- list(
  dwellings_latest = "https://www.abs.gov.au/statistics/economy/price-indexes-and-inflation/total-value-dwellings/latest-release",
  dwellings_workbook = "https://www.abs.gov.au/statistics/economy/price-indexes-and-inflation/total-value-dwellings/mar-quarter-2026/643201.xlsx",
  lending_latest = "https://www.abs.gov.au/statistics/economy/finance/lending-indicators/latest-release",
  approvals_latest = "https://www.abs.gov.au/statistics/industry/building-and-construction/building-approvals-australia/latest-release",
  home_ownership_support = "https://treasury.gov.au/policy-topics/housing/home-ownership-support"
)
```

## Data

The main source is ABS [Total Value of Dwellings](https://www.abs.gov.au/statistics/economy/price-indexes-and-inflation/total-value-dwellings/latest-release), March quarter 2026. The release estimates the total value, number and mean price of Australia's residential dwellings. I also keep [Lending Indicators](https://www.abs.gov.au/statistics/economy/finance/lending-indicators/latest-release), [Building Approvals](https://www.abs.gov.au/statistics/industry/building-and-construction/building-approvals-australia/latest-release), and Treasury's [home ownership support](https://treasury.gov.au/policy-topics/housing/home-ownership-support) page in view because prices alone do not tell us whether buyers are being helped, stretched, or both.

```{r}
#| label: load-data

download_if_missing <- function(url, path) {
  if (!file.exists(path)) {
    download.file(url, path, mode = "wb", quiet = TRUE)
  }
  path
}

as_number <- function(x) {
  parse_number(as.character(x), na = c("", "NA", "n/a", "-"))
}

quarter_date_from_excel <- function(x) {
  as.Date(as.numeric(x), origin = "1899-12-30")
}

state_from_item <- function(item) {
  str_split_fixed(item, ";", 3)[, 2] |>
    str_squish()
}

measure_from_item <- function(item) {
  case_when(
    str_detect(item, "Value of dwelling stock; Owned by All Sectors") ~ "Total dwelling stock value",
    str_detect(item, "Mean price of residential dwellings") ~ "Mean dwelling price",
    str_detect(item, "Number of residential dwellings") ~ "Number of dwellings",
    TRUE ~ NA_character_
  )
}

workbook_path <- download_if_missing(
  source_urls$dwellings_workbook,
  file.path(raw_dir, "abs-total-value-dwellings-mar-quarter-2026.xlsx")
)

raw <- read_excel(workbook_path, sheet = "Data1", col_names = FALSE)

series_meta <- tibble(
  column = seq_along(raw),
  item = as.character(unlist(raw[1, ])),
  unit = as.character(unlist(raw[2, ])),
  measure = measure_from_item(item),
  state = state_from_item(item)
) |>
  filter(!is.na(measure), !is.na(state), state != "")

dwelling_series <- raw |>
  slice(-(1:10)) |>
  transmute(row_id = row_number(), quarter_date = quarter_date_from_excel(...1)) |>
  bind_cols(raw |> slice(-(1:10)) |> select(all_of(series_meta$column))) |>
  pivot_longer(-c(row_id, quarter_date), names_to = "column_name", values_to = "value") |>
  mutate(
    column = as.integer(str_remove(column_name, "\\.\\.\\.")),
    value = as_number(value)
  ) |>
  left_join(series_meta, by = "column") |>
  filter(!is.na(value), !is.na(quarter_date), !is.na(measure)) |>
  mutate(
    value = case_when(
      measure == "Total dwelling stock value" ~ value / 1000,
      measure == "Mean dwelling price" ~ value * 1000,
      measure == "Number of dwellings" ~ value * 1000,
      TRUE ~ value
    )
  )

latest_quarter <- max(dwelling_series$quarter_date, na.rm = TRUE)

latest_state_prices <- dwelling_series |>
  filter(measure == "Mean dwelling price", quarter_date %in% sort(unique(quarter_date), decreasing = TRUE)[1:2], state != "Australia") |>
  select(quarter_date, state, mean_price = value) |>
  pivot_wider(names_from = quarter_date, values_from = mean_price) |>
  rename(previous_price = 2, latest_price = 3) |>
  mutate(
    quarterly_change = latest_price / previous_price - 1,
    dollar_change = latest_price - previous_price
  )
```

## What Has Happened Nationally?

```{r}
#| label: national-stock-value
#| fig-cap: "ABS total value of Australian residential dwelling stock."

dwelling_series |>
  filter(measure == "Total dwelling stock value", state == "Australia", quarter_date >= as.Date("2021-03-31")) |>
  ggplot(aes(quarter_date, value)) +
  geom_line(linewidth = 0.9, colour = "#126782") +
  geom_point(size = 2, colour = "#126782") +
  scale_y_continuous(labels = dollar_format(suffix = "t", scale = 1 / 1000)) +
  labs(x = NULL, y = "Total dwelling stock value")
```

```{r}
#| label: national-mean-price
#| fig-cap: "ABS mean dwelling price for Australia."

dwelling_series |>
  filter(measure == "Mean dwelling price", state == "Australia", quarter_date >= as.Date("2021-03-31")) |>
  ggplot(aes(quarter_date, value)) +
  geom_line(linewidth = 0.9, colour = "#7a4b9d") +
  geom_point(size = 2, colour = "#7a4b9d") +
  scale_y_continuous(labels = dollar) +
  labs(x = NULL, y = "Mean dwelling price")
```

The March quarter 2026 release puts the national mean dwelling price at over $1.1 million. That is not a typical transaction price and it is not an affordability measure, but it is a useful whole-market signal: the stock keeps getting more expensive, even before we ask who can actually buy into it.

## Which States Moved Most Recently?

```{r}
#| label: latest-state-prices
#| fig-cap: "Latest ABS mean dwelling prices by state and territory."

latest_state_prices |>
  ggplot(aes(fct_reorder(state, latest_price), latest_price, fill = quarterly_change)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = dollar) +
  scale_fill_gradient2(labels = percent, low = "#8c2d04", mid = "grey80", high = "#006d2c", midpoint = 0) +
  labs(x = NULL, y = "Mean dwelling price", fill = "Quarterly change")
```

```{r}
#| label: latest-state-table

latest_state_prices |>
  arrange(desc(quarterly_change)) |>
  transmute(
    state,
    latest_mean_price = dollar(latest_price, accuracy = 100),
    quarterly_change = percent(quarterly_change, accuracy = 0.1),
    dollar_change = dollar(dollar_change, accuracy = 100)
  ) |>
  kable()
```

## Policy Questions This Sets Up

Treasury says the 5% Deposit Scheme was expanded from 1 October 2025, with uncapped places, removed income caps and higher property price caps. That creates a nice empirical question, but not a simple one.

The questions I would ask next:

- Did first-home buyer loan commitments rise after the deposit-scheme expansion, and was that concentrated in states where prices were already rising fastest?
- Did the policy mostly pull forward demand, or did it change the composition of buyers entering the market?
- Are building approvals moving in the same places as dwelling prices, or is supply still lagging the demand signal?
- Are investor loans and first-home buyer loans moving together or pushing against each other?
- If mean prices rise after access support expands, how much of the buyer benefit is eaten by price pressure?

Those need lending and building-approval data joined to this price series. This post is just the first brick.

## Caveats

- Mean dwelling price is a stock estimate, not a median sale price or a first-home buyer price.
- State averages hide massive variation inside cities, regions and property types.
- The latest ABS dwelling estimates are preliminary and can be revised.
- Prices do not measure affordability on their own. We need income, rents, interest rates, deposit constraints, lending standards and supply.
- Policy timing matters. A demand-side policy can show up in lending before it shows up in prices, and supply responses can be much slower.
